### 2025-07-08 factor-analysis.qmd
20250228 - remove duplicate label
20250618 - fix formulas
20250707 - updates

```{r}

```

@@

$$
\bar{X} = \lambda \cdot \text{F} + \text{Item Intercept} + \text{Error Term}
$$ {#eq-indicatorLatentAssociation}

@@

## Exploratory Factor Analysis {#sec-cfa}

@@

@@

>>>>

```{r}
library("psych")
library("nFactors")
library("lavaan")
library("lavaangui")
library("tidyverse")
```

### Load Data {#sec-factorAnalysisLoadData}

```{r}
#| eval: false
#| include: false

load(file = file.path(path, "/OneDrive - University of Iowa/Teaching/Courses/Fantasy Football/Data/player_stats_weekly.RData", fsep = ""))
load(file = file.path(path, "/OneDrive - University of Iowa/Teaching/Courses/Fantasy Football/Data/nfl_nextGenStats_weekly.RData", fsep = ""))
```

```{r}
load(file = "./data/player_stats_weekly.RData")
load(file = "./data/nfl_nextGenStats_weekly.RData")
```

### Prepare Data {#sec-factorAnalysisPrepareData}

#### Merge Data {#sec-factorAnalysisPrepareDataMerge}

```{r}
dataMerged <- full_join(
  player_stats_weekly,
  nfl_nextGenStats_weekly %>% select(-any_of(c("player_display_name","completions","attempts","receptions","targets"))),
  by = c("player_id" = "player_gsis_id","season","season_type","week"),
)
```

#### Specify Variables {#sec-pcaPrepareDataSpecifyVars}

```{r}
faVars <- c(
  "completions","attempts","passing_yards","passing_tds","passing_interceptions",
  "sacks_suffered","sack_yards_lost","sack_fumbles","sack_fumbles_lost",
  "passing_air_yards","passing_yards_after_catch","passing_first_downs",
  "passing_epa","passing_cpoe","passing_2pt_conversions","pacr","pass_40_yds",
  "pass_inc","pass_comp_pct","fumbles","two_pts",
  "avg_time_to_throw","avg_completed_air_yards","avg_intended_air_yards",
  "avg_air_yards_differential","aggressiveness","max_completed_air_distance",
  "avg_air_yards_to_sticks","passer_rating", #,"completion_percentage"
  "expected_completion_percentage","completion_percentage_above_expectation",
  "avg_air_distance","max_air_distance")
```

@@

$$
X = \lambda \cdot \text{F} + \text{Item Intercept} + \text{Error Term}
$$ {#eq-indicatorLatentAssociation}

@@

## Exploratory Factor Analysis {#sec-efa}

@@

### Example {#sec-efaExample}

We generated the scree plot in @fig-efaScreePlot1 using the `fa.parallel()` function of the `psych` package [@R-psych].

```{r}
#| label: fig-efaScreePlot1
#| fig-cap: "Scree Plot."
#| fig-alt: "Scree Plot."

psych::fa.parallel(
  x = dataMerged[faVars],
  fm = "minres", # errors out when using "ml"
  fa = "fa"
)
```

We generated the scree plot in @fig-efaScreePlot2 using the `nScree()` and `plotnScree()` functions of the `nFactors` package [@R-nFactors].

```{r}
#| label: fig-efaScreePlot2
#| fig-cap: "Scree Plot."
#| fig-alt: "Scree Plot."

screeDataEFA <- nFactors::nScree(
  x = cor(
    dataMerged[faVars],
    use = "pairwise.complete.obs"),
  model = "factors")

nFactors::plotnScree(screeDataEFA)
```

We generated the very simple structure (VSS) plots in Figures [-@fig-efaVSSPlot1] and [-@fig-efaVSSPlot2] using the `vss()` and `nfactors()` functions of the `psych` package [@R-psych].

```{r}
#| label: fig-efaVSSPlot1
#| fig-cap: "Very Simple Structure Plot."
#| fig-alt: "Very Simple Structure Plot."

psych::vss(
  dataMerged[faVars],
  rotate = "oblimin",
  fm = "minres") # errors out when using "mle"
```

```{r}
#| label: fig-efaVSSPlot2
#| fig-cap: "Very Simple Structure Plot."
#| fig-alt: "Very Simple Structure Plot."

psych::nfactors(
  dataMerged[faVars],
  rotate = "oblimin",
  fm = "minres") # errors out when using "ml"
```

We fit EFA models using the `efa()` function of the `lavaan` package [@Rosseel2012_packages; @R-lavaan].

```{r}
#| eval: false

efaOblique_fit <- lavaan::efa(
  data = dataMerged,
  ov.names = faVars,
  nfactors = 1:9,
  rotation = "geomin",
  missing = "ML",
  estimator = "MLR",
  meanstructure = TRUE
)
```

A path diagram of the EFA model was created using the `plot_lavaan()` function of the `lavaangui` package [@R-lavaangui; @Karch2025_packages].

```{r}
#| eval: false

lavaangui::plot_lavaan(efaOblique_fit)
```

@@

### Example {#sec-cfaExample}

We fit CFA models using the `cfa()` function of the `lavaan` package [@Rosseel2012_packages; @R-lavaan].

```{r}
cfaModel1_syntax <- '
 #Factor loadings
 F1 =~ completions + attempts + passing_yards + passing_tds + passing_interceptions + sacks_suffered + sack_yards_lost + sack_fumbles + sack_fumbles_lost + passing_air_yards + passing_yards_after_catch + passing_first_downs + pass_inc + fumbles
 F2 =~ passing_air_yards + pacr + avg_completed_air_yards + avg_intended_air_yards + aggressiveness + max_completed_air_distance + avg_air_yards_to_sticks + expected_completion_percentage + avg_air_distance + max_air_distance
 F3 =~ passing_tds + passing_epa + passing_cpoe + pass_40_yds + pass_comp_pct + avg_air_yards_differential + max_completed_air_distance + avg_time_to_throw + avg_completed_air_yards + avg_intended_air_yards + passer_rating + completion_percentage_above_expectation
'
```

```{r}
#| eval: false

cfaModel1_fit <- lavaan::cfa(
  cfaModel1_syntax,
  data = dataMerged,
  missing = "ML",
  estimator = "MLR",
  std.lv = TRUE)
```

```{r}
#| eval: false

summary(
  cfaModel1_fit,
  fit.measures = TRUE,
  standardized = TRUE,
  rsquare = TRUE)
```

```{r}
#| eval: false

lavaan::fitMeasures(
  cfaModel1_fit,
  fit.measures = c(
    "chisq", "df", "pvalue",
    "chisq.scaled", "df.scaled", "pvalue.scaled",
    "chisq.scaling.factor",
    "baseline.chisq","baseline.df","baseline.pvalue",
    "rmsea", "cfi", "tli", "srmr",
    "rmsea.robust", "cfi.robust", "tli.robust"))
```

```{r}
#| eval: false

lavaan::residuals(
  cfaModel1_fit,
  type = "cor")
```

```{r}
#| eval: false

lavaan::modificationindices(
  cfaModel1_fit,
  sort. = TRUE)
```

```{r}
#| eval: false

lavaan::lavPredict(cfaModel1_fit)
```

A path diagram of the CFA model was created using the `plot_lavaan()` function of the `lavaangui` package [@R-lavaangui; @Karch2025_packages].

```{r}
#| eval: false

lavaangui::plot_lavaan(cfaModel1_fit)
```

### 2025-07-08 index.qmd
20250707 - updates

# Automatically create a bib database for R packages
packagesForBibliography <- c(
  "petersenlab", "ffanalytics",
  "knitr", "rmarkdown", "dplyr", "ggplot2", "tidyverse", "broom", "broom.mixed", 
  "downlit", "xml2", "nflreadr", "nflplotR", "nflfastR", "nflseedR", "nfl4th", 
  "gsisdecoder", "progressr", "DescTools", "pwr", "pwrss", "WebPower", "XICOR", 
  "dagitty", "ggdag", "lubridate", "ggtext", "gghighlight", "ggExtra", "grid", 
  "patchwork", "pROC",  "lme4", "lmerTest", "MuMIn", "emmeans", "pbkrtest", 
  "sjstats", "AICcmodavg", "rstan", "brms", "tidybayes", "bbmle", 
  "fitdistrplus", "sn", "mclust", "magrittr", "viridis", "plotly", "quantmod", 
  "fPortfolio", "NMOF", "nFactors", "xts", "zoo", "forecast", "stringi", #"webshot2"
  "parallel", "parallelly", "doParallel", "missRanger", "ggridges", "powerjoin", 
  "caret", "tidymodels", "LongituRF", "gpboost", "cmdstanr", "learnr", 
  "remotes", "corrplot", "correlation", "mgcv", "rms", "car", "performance", 
  "lavaan", "mice", "miceadds", "interactions", "shiny", "PearsonDS", 
  "robustbase", "ordinal", "MASS", "data.table", "future", "future.apply", 
  "QuantPsyc", "SimDesign", "effectsize"
)

>>>>

# Automatically create a bib database for R packages
packagesForBibliography <- c(
  "petersenlab", "ffanalytics",
  "knitr", "rmarkdown", "dplyr", "ggplot2", "tidyverse", "broom", "broom.mixed", 
  "downlit", "xml2", "nflreadr", "nflplotR", "nflfastR", "nflseedR", "nfl4th", 
  "gsisdecoder", "progressr", "DescTools", "pwr", "pwrss", "WebPower", "XICOR", 
  "dagitty", "ggdag", "lubridate", "ggtext", "gghighlight", "ggExtra", "grid", 
  "patchwork", "pROC",  "lme4", "lmerTest", "MuMIn", "emmeans", "pbkrtest", 
  "sjstats", "AICcmodavg", "rstan", "brms", "tidybayes", "bbmle", 
  "fitdistrplus", "sn", "mclust", "magrittr", "viridis", "plotly", "quantmod", 
  "fPortfolio", "NMOF", "nFactors", "xts", "zoo", "forecast", "stringi", #"webshot2"
  "parallel", "parallelly", "doParallel", "missRanger", "ggridges", "powerjoin", 
  "caret", "tidymodels", "LongituRF", "gpboost", "cmdstanr", "learnr", 
  "remotes", "corrplot", "correlation", "mgcv", "rms", "car", "performance", 
  "lavaan", "lavaangui", "mice", "miceadds", "interactions", "shiny", 
  "PearsonDS", "robustbase", "ordinal", "MASS", "data.table", "future", 
  "future.apply", "QuantPsyc", "SimDesign", "effectsize"
)
